---
title: "Relationships in Data"
subtitle: "Week 5"
date: 2025-10-17
format: 
  html:
    embed-resources: true
toc: true
---

# Before we start

-   Any questions?


<center>
<a href="https://artur-baranov.github.io/nu-ps403-ds/scripts/ps403-d_05.qmd" class="btn btn-primary" role="button" download="ps403-d_05.qmd" style="width:200px" target="_blank">Download script</a>
</center>
<br>
<center>
<a href="https://artur-baranov.github.io/nu-ps403-ds/data/whogov.csv" class="btn btn-primary" role="button" download="whogov.csv" style="width:200px" target="_blank">Download data</a>
</center><br>


whogov: create a variable independent/partisan for the review

::: callout-note
## Review

First, load the `tidyverse` library

```{r}
#| message: false

library(tidyverse)
```

Load the WhoGov dataset.

```{r}
whogov = read.csv("data/WhoGov.csv")
```

First of all, **data wrangling** (*take a look on week 2 script*).

Create a variable for independent and partisan leaders using `case_when()`.

```{r}
whogov = whogov %>%
  mutate(indep = case_when(leader_party == "independent" ~ "Independent",
                           leader_party != "indepdenent" ~ "Partisan"))
```

Check the distribution of the newly created variable using `table()`

```{r}
table(whogov$indep)
```

```{r}
ggplot(whogov) +
  
```


```{r}
ggplot(whogov %>% filter(year >= 2000,
                         system_category %in% c("Military dictatorship",
                                                "Parliamentary democracy"))) +
  geom_boxplot(aes(x = system_category,
                   y = average_total)) +
  facet_grid(~ indep)
```

```{r}
whogov %>%
  t_test(n_individuals ~ indep)
```



:::


# Agenda

-   Comparing distributions, *statistically*

-   Analyzing relationships, *statistically*


ggally?

t.test and regressions