---
title: "Data Management with GitHub"
subtitle: "Week 6"
date: 2025-10-24
format: html
engine: knitr
filters:
  - webr
webr:
  packages:
    - dplyr
    - knitr
toc: true
---

# Before we start

-   Any questions?

-   Today the class is different: we will cover the review on the Website.

::: callout-note
## Review

Let's quickly review joins in R. Exercises are adapted from [here](https://www.r-bloggers.com/2022/01/how-to-join-tables-in-r).

Create a sample dataset with information on how far the students live from campus. Why are we using `set.seed()`? Do you remember what `runif()` function does?

```{webr-r}
#| results: 'asis'
set.seed(123)
student_distance = data.frame(student = seq(1, 6),         
                              distance = runif(6, 3, 10))

head(student_distance)
```

Create another data frame with information on how students get to school.

```{webr-r}
#| results: 'asis'
student_transport = data.frame(Student = seq(1, 7, by = 2),
                               Transport = c("Bus", "Carpool", "Walk", "Bus"))

head(student_transport) %>%
  kable("html")
```

First, take `student_distance` dataset and using `left_join()` merge the `student_transport` dataframe.

```{webr-r}
#| results: 'asis'

student_distance %>%
  left_join(..., by = ...) 
```

Now, use `full_join()`. What is the difference?

```{webr-r}
#| results: 'asis'

...
```

Take `student_transport` and use `left_join()` to merge the `student_distance`. What is the difference? How would you make a choice?

```{webr-r}
#| results: 'asis'

...
```


:::

# Agenda

-   Working with GitHub

-   Covering extra topics (if time allows!)

# GitHub

You already know what GitHub is! But GitHub desktop is a bit more intuitive than working with the Git within RStudio terminal. 

Download GitHub Desktop. Then, clone the repository we are working with Today. Let's do it together, step-by-step.

<center>
<a href="https://desktop.github.com/download" class="btn btn-primary" style="width:200px" target="_blank">GitHub Desktop</a>
</center>
<br>
<center>
<a href="https://github.com/artur-baranov/github-desktop" class="btn btn-primary" style="width:200px" target="_blank">Repository</a>
</center>


# Check List

- [ ] I understand what repository is, what fork mean, and what the Git is

- [ ] I know how to use GitHub desktop to create and clone repositories

- [ ] I can use `patchwork` library to easily plot the graphs on top of each another

- [ ] Interpretation of boxplot does not scare me!

# Optional Exercises

Download this script.

<center>
<a href="https://artur-baranov.github.io/nu-ps403-ds/scripts/ps403-d_06.qmd" class="btn btn-primary" role="button" download="ps403-d_06.qmd" style="width:200px" target="_blank">Download script</a>
</center><br>

Load the `tidyverse` and `patchwork` libraries.

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

...
```

:::

Move the `CSES_MARPOR.rds` that you downloaded from GitHub to your data folder. Load it into R.

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

cses = readRDS("data/CSES_MARPOR.rds")
```

:::

Draw a histogram of `opportunity` variable. Save it to `p1` object.

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

p1 = ...
```

:::

Draw a histogram of `loyalty` variable. Save it to `p2` object.

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

p2 = ...
```

:::

Display both graphs (`p1` and `p2`) side-by-side using `+` operator from `patchwork` library.

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

... + p2
```

:::

Customize the graphs (add axes names and title, and change theme to `theme_bw()`), and display two graphs together again.

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

...
```

:::