---
title: "Data Presentation Using Tables"
subtitle: "Week 3"
date: 2025-10-03
format: 
  html:
    embed-resources: true
toc: true
---

# Before we start

-   Any questions?

-   For sampling hints for the next homework lab, look for Week 1 script.

<center>
<a href="https://artur-baranov.github.io/nu-ps403-ds/scripts/ps403-d_03.qmd" class="btn btn-primary" role="button" download="ps403-d_03.qmd" style="width:200px" target="_blank">Download script</a>
</center>
<br>
<center>
<a href="https://artur-baranov.github.io/nu-ps403-ds/data/wvs.csv" class="btn btn-primary" role="button" download="wvs.csv" style="width:200px" target="_blank">Download data</a>
</center><br>

::: callout-note
## Review

First, load the `tidyverse` library

```{r}
#| message: false

library(tidyverse)
```

Then, load the V-Dem data

```{r}
#| eval: false

load(url("https://github.com/vdeminstitute/vdemdata/raw/6bee8e170578fe8ccdc1414ae239c5e870996bc0/data/vdem.RData"))
```


You are interested in corruption in various geographical regions. Select year (`year`), region (`e_regionpol_7C`) and corruption (`v2x_corr`) variables from the `vdem` dataset.

```{r}
#| eval: false

corruption_data = ... %>%
  ...(year, ..., v2x_corr) 
```

As we are working with the V-Dem data, let's rename variables so it's more straightforward.

```{r}
#| eval: false

corruption_data = corruption_data ...
  ...(region = e_regionpol_7C,
      corruption = ...)
```

Using `group_by()` and `summarize()` calculate the average corruption for all regions in the dataset.

```{r}
#| eval: false

corruption_data ...
  (...) %>%
  ...(average = ...(..., na.rm = T))
```

Lastly, let's see the distributions of corruption across regions. Draw a boxplot below with `region` variable on the X axis, and `corruption` variable on the Y axis.

```{r}
#| eval: false

ggplot(data = corruption_data) +
  ...(aes(x = ..., y = ...))
```

Something is wrong, right? We haven't checked the classes of the variables, and apparently the `region` variable has to be changed. Let's check it's class first.

```{r}
#| eval: false

...(corruption_data$...)
```

What class should it be? Let's recode it directly on the plot.

```{r}
#| eval: false

...(...) +
  geom_boxplot(aes(x = ...(region), y = ...))
```

Finally, remove the data we worked with from the environment.

```{r}
#| eval: false

rm(vdem, corruption_data)
```


:::


# Agenda

-   Summarizing Data

-   Presenting Data with Tables

# Exploring Data

Today we are working with the World Values Survey data. You can find the codebook [here](https://www.worldvaluessurvey.org/WVSDocumentationWV7.jsp).

Load the data. Explore the dimensions. This is a smaller version of the dataset, the full one is available on the WVS [website](https://www.worldvaluessurvey.org/wvs.jsp).

```{r}
#| results: hold

wvs = read.csv("data/wvs.csv")

ncol(wvs)
nrow(wvs)
```

Explore the names of the variables.

```{r}
colnames(wvs)
```

For your convenience, here are the variables' description from the codebook:

-   `B_COUNTRY_ALPHA` - country name

-   `Q108` - Government´s vs individual´s responsibility (10 - people should take more responsibility to provide for themselves, 1 - the government should take more responsibility to ensure that everyone is provided for)

-   `Q191` - Justifiable: Violence against other people (10 - justifiable, 1 - not justifiable)

-   `Q194` - Justifiable: Political violence (10 - justifiable, 1 - not justifiable)

Let's get the **range** of data. What does it mean? 

```{r}
range(wvs$Q108)
```

Let's double check with summary. Same! Take a look on the codebook.

```{r}
summary(wvs$Q108)
```

For the purposes of this exercise, drop all missing or irrelevant answers (which you shoudl be **very** careful erasing in real task).

```{r}
wvs = wvs %>%
  filter(Q108 > 0)
```

And for our convenience, rename the variables.

```{r}
wvs = wvs %>%
  rename(country = B_COUNTRY_ALPHA,
         responsibility = Q108,
         violence = Q191,
         political_violence = Q194)
```

# Presenting Data in a Table

Now, let's present the descriptive statistics for the `responsibility` variable, but for each country. Again, tricky `group_by()` command! Make sure to include `mean()`, `sd()` and `median()`.

```{r}
responsibility_description = wvs %>%
  group_by(country) %>%
  summarize(average = mean(responsibility),
            sd = sd(responsibility),
            median = median(responsibility))

responsibility_description %>%
  head()
```

Great! Now, let's make publishable table to present the variable. However, focus only on Post-Soviet countries. Make sure to go through syntax!

```{r}
post_soviet = c("ARM", "AZE", "BLR", "EST", "GEO", "KAZ", "KGZ", "LVA", "LTU", "MDA", "RUS", "TJK",  "TKM", "UKR", "UZB")

post_soviet_responsibility = responsibility_description %>%
  filter(country %in% post_soviet)

post_soviet_responsibility
```

To make it publishable and nice-looking load the `tinytable` library.

```{r}
#| message: false

library(tinytable)
```

Try it out! Looks great, but we can do more.

```{r}
post_soviet_responsibility %>%
  tt()
```

Make sure to format it well. For example, capitalize the names of the columns. Other functions for changing the character case [here](https://stringr.tidyverse.org/reference/case.html) - you can apply it to observations, too.

```{r}
colnames(post_soviet_responsibility) = colnames(post_soviet_responsibility) %>%
  str_to_title() 
```

Now, add caption, and round up numbers. Highlight, Central Asian countries: Kazakhstan (KAZ), Kyrgyzstan (KGZ), Tajikistan (TJK) and Uzbekistan (UZB).

```{r}
post_soviet_responsibility %>%
  tt(caption = "Descriptive Statistics for Post Soviet Region") %>%
  format_tt(digits = 2) %>%
  style_tt(i = c(2, 3, 5, 6), # rows
           j = 1:4, # columns
           background = "teal",
           color = "white",
           bold = TRUE)
```

# Extra

Here's another example of what you can do with this. First, prepare the data. Then, draw a `tinytable`. But this is **extra**!

```{r}
post_soviet_description = wvs %>%
  group_by(country) %>%
  summarize(average_r = mean(responsibility),
            sd_r = sd(responsibility),
            average_v = mean(violence),
            sd_v = sd(violence),
            average_pv = mean(political_violence),
            sd_pv = sd(political_violence)) %>%
  filter(country %in% post_soviet) 

extra_table = post_soviet_description %>%
  tt() %>%
  group_tt(j = list("Responsibility" = 2:3, 
                    "Violence" = 4:5,
                    "Political Violence" = 6:7)) %>%
  { colnames(.) = c("Country", "Average", "SD", "Average", "SD", "Average", "SD"); . } %>%
  format_tt(digits = 2)

extra_table
```

You even can add plots into the table! You can explore it [here](https://vincentarelbundock.github.io/tinytable).

# Check list

- [ ] I am confident in summarizing variables using combination of `group_by()` and `summarize()`

- [ ] I can present nice looking tables with `tinytable`

- [ ] I know how to use `%in%` operator to filter data

# Optional Exercises

Explore `violence` variable in `wvs` dataset. What is its range?

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

...
```

:::

Draw a histogram of `violence` variable using `ggplot()`.

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

...
```

:::

Anything strange? What does negative values stand for? Remove if needed.

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

...
```

:::

Present average and median for `violence` for each country.

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

...
```

:::

Now, present the data for Latin American countries using `tinytable`. Here is the vector for your convenience.

```{r}
latin_america <- c(
  "ARG",  # Argentina
  "BOL",  # Bolivia
  "BRA",  # Brazil
  "CHL",  # Chile
  "COL",  # Colombia
  "CRI",  # Costa Rica
  "CUB",  # Cuba
  "DOM",  # Dominican Republic
  "ECU",  # Ecuador
  "SLV",  # El Salvador
  "GTM",  # Guatemala
  "HND",  # Honduras
  "MEX",  # Mexico
  "NIC",  # Nicaragua
  "PAN",  # Panama
  "PRY",  # Paraguay
  "PER",  # Peru
  "URY",  # Uruguay
  "VEN"   # Venezuela
)

```

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

...
```

:::

Customize the `tinytable` so the columns look like titles. Highlight top-3 most violent countries with `red` color.

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

...
```

:::
