---
title: "Data Wrangling and Classes"
subtitle: "Week 2"
date: 2025-09-26
format: 
  html:
    embed-resources: true
toc: true
---

# Before we start

-   Any questions?

<center>
<a href="https://artur-baranov.github.io/nu-ps403-ds/scripts/ps403-d_02.qmd" class="btn btn-primary" role="button" download="ps403-d_01.qmd" style="width:200px" target="_blank">Download script</a>
</center>

# Review of the Previous Session

Let's refresh what we did last week. 

::: callout-note
## Review

First of all, let's load the `tidyverse` and `readxl` libraries. Then, load the Comparative Political Dataset from the last class.

```{r}
library(tidyverse)
library(readxl)

cpds = read_excel("data/cpds.xlsx")
```

Imagine you are interested in fractionalizon of electoral systems (you can read more about fractionalization after class [here](https://sk.sagepub.com/ency/edvol/the-encyclopedia-of-political-science/chpt/fractionalization-index)). Subset the following variables from the dataset: `year`, `country`, `rae_ele` and `fed`.

```{r}
#| eval: false

cpds_subset = cpds ...
  ...(year, ...)
```

Rename the `rae_ele` to `fractionalization` in the `cpds_subset` dataframe.

```{r}
#| eval: false

... = cpds_subset %>%
  ...(fractionalization = ...)
```

Draw a histogram to show the distribution of Douglas Rae's index of electoral `fractionalization.`

```{r}
#| eval: false

... +
  geom_histogram(...)
```

What is the minimum value? What is the maximum value? 

```{r}
#| eval: false

...
```

The index indicates that 1 corresponds to a fragmented system (votes are widely dispersed across parties), while 0 means that everyone voted for the same party. What is the average?

```{r}

```

Draw a `geom_boxplot()` to see the relationship between the federalism (`fed`) on the x axis and `fractionalization` on the y axis. The federalism is coded as: 0 = no; 1 = weak; 2 = strong. Is something wrong? 

```{r}
#| eval: false

...
  ...(aes(...))
```

Don't forget to remove the `cpds` and `cpds_subset` dataframes.

```{r}
#| eval: false

rm(cpds, cpds_subset)
```

:::

# Agenda for Today

-   Data Classes

-   Data Wrangling

-   Presenting data with `tinytable`

# Exploring and Wrangling Data

Let's load the data

```{r}
load(url("https://github.com/vdeminstitute/vdemdata/raw/6bee8e170578fe8ccdc1414ae239c5e870996bc0/data/vdem.RData"))
```

This is the V-Dem dataset. For your reference, their codebook is available [here](https://v-dem.net/documents/38/V-Dem_Codebook_v14.pdf).

The dataset is huge! Be careful

```{r}
#| results: hold
nrow(vdem)
ncol(vdem)
```

Imagine you are interested in the relationship between regime type and physical violence. Let’s `select()` and rename the variables we will work with. Quite unfortunately, the names of the variables are not as straightforward. The regime index is `e_v2x_polyarchy_5C` and Physical violence index is `v2x_clphy`. And you can combine both actions in the same function!

```{r}
violence_data = vdem %>%
  select(country_name, 
         year, 
         regime = e_v2x_polyarchy_5C, 
         violence = v2x_clphy) 
```

Let's see the result

```{r}
head(violence_data)
```

That was the first step. Let's subset even more data. Let's analyze the distribution of the Violence Index in the year 2000. We need to **filter** the data for this task.

```{r}
violence_2000 = violence_data %>%
  filter(year == 2000)
```

The higher the number, the better the physical integrity in a given country. 

```{r}
#| warning: false

ggplot(data = violence_2000) +
  geom_histogram(aes(x = violence))
```

Let's customize the plot a bit.

```{r}
ggplot(data = violence_2000) +
  geom_histogram(aes(x = violence)) +
  labs(title = "Physical Integrity Rights Index",
       subtitle = "In 2000",
       x = "Violence Integrity Index",
       y = "") +
  theme_bw()
``` 

Imagine, you asked the following question. Is it true that in the year 2000 democracies had better violence integrity rights index?

Firstly, you we need to differentiate between the regimes. Using `mutate()` function we can create and modify existing variables. Let's do that!

```{r}
violence_2000 = violence_2000 %>%
  mutate(democracy = case_when(regime >= 0.5 ~ "Democracy",
                               regime < 0.5 ~ "Autocracy"))

head(violence_2000)
```

There are multiple ways to visually compare between two groups. 

Let's draw histograms first.

```{r}
#| message: false

ggplot(violence_2000) +
  geom_histogram(aes(x = violence, fill = democracy))
```

But comparing distributions between groups is easier with the boxplot. Are they different?

```{r}
ggplot(violence_2000) +
  geom_boxplot(aes(x = democracy, y = violence))
```

# Data Classes

You can represent the same concepts with numbers or characters. For example, we have two variables indicating the regime type. One of which is more granular, `regime`. Another is binary, which is `democracy`. Explore the latter first.

```{r}
with(violence_2000,
     table(regime))
```

Now, explore the former. Side note, **pay attention to syntax**, the functions are almost equivalent. 

```{r}
table(violence_2000$democracy)
```
Theoretically, we can plot even more detailed graph using `geom_boxplot()` if we use `regime` data. Let's see. Something fails! 

```{r}
ggplot(violence_2000) +
  geom_boxplot(aes(x = regime, y = violence))
```

Let's correct. Now it's better!

```{r}
ggplot(violence_2000) +
  geom_boxplot(aes(x = as.factor(regime), y = violence))
```

The problem is, by default R treats `regime` variable as continuous, but it's a categorical! Explore the graph below.

```{r}
library(DiagrammeR)
mermaid("
graph LR
    D[Data] --> C[Categorical]
    D --> N[Numerical]
    C --> no[Nominal]
    C --> Or[Ordinal]
    N --> di[Discrete]
    N --> co[Continuous]
    no --> c[Character]
    Or --> f[Factor]
    di --> i[Integer]
    co --> n[Numeric]
")
```

You can check the class using `class()` function.

```{r}
class(violence_2000$regime)
```

And you can make it a category by transforming it to factor using `as.factor()`. Other functions are `as.numeric()`and `as.character()`.

```{r}
violence_2000$regime = as.factor(violence_2000$regime)
```

# Summarizing and Presenting data 

Treating a variable as a categorcal would help us to summarize data per category. For example, you want to present information of average level of violence in different regimes. 

Let's switch to the main dataset. Quickly prepare the data using `case_when()` function -- a better version of `ifelse()`. Here we are using the codebook's guidence to rename the categories.

```{r}
#| warning: false

violence_data = violence_data %>%
  mutate(regime = case_when(regime > 0.75 ~ "Democratic",
            regime == 0.75 ~ "Minimally Democratic",
            regime == 0.5 ~ "Ambivalent",
            regime == 0.25 ~ "Autocratic",
            regime < 0.25 ~ "Closed Autocratic"))

head(violence_data)
```

A tricky but helpful operation is groupping a variable, and then calculating statistics. [This picture really helps me wrap my head around it](https://artur-baranov.github.io/images/groupby_sum_eng.png).

```{r}
regime_violence = violence_2000 %>%
  group_by(regime) %>%
  summarize(average = mean(violence))

regime_violence
```

Finally, let's present that in a nice way using `tinytable` library.

```{r}
library(tinytable)

regime_violence %>% 
  tt(title = "Average level of violence in different regimes")
```

# Check list

- [ ] I am confident in drawing distributions: either using `geom_histogram()` or `geom_boxplot()`

- [ ] I know what data classes are, and how to change them, too.

- [ ] I know how to recode variable using combination of `case_when()` and `mutate()`

- [ ] I can easily `summarize()` variables using `group_by()` to present statistics by category

- [ ] I can present nice tables with `tinytable` library

# Optional Exercises

Create a new `violence_2010` dataset, by subsetting `violence_data` for year 2010.

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

violence_2010 = violence_data %>%
  ...
```

:::

Using `mutate()` and `case_when()`, create a new variable `violence_binary` that is "High integrity" if the violence variable is greater or equal to 0.5, and "Low integrity" if it is less than 0.5.

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

violence_2010 = violence_2010 %>%
  ...(violence_binary = case_when(... >= 0.5 ~ "High Integrity",
                                     ...))
```

:::

Compare the two variables coding: `violence` and `violence_binary`. Check there classes. What is the difference? Briefly describe the type of each variable. Imagine you were to draw a boxplot, how would class differences in the variables define their use case on the graph?

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}
#| eval: false

...
```

:::

Let's calculate the average regime type per `violence_binary`. Use `group_by()` and `summarize()`. Save it to `output` object.

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

output = violence_2010 %>%
  ...
```

:::

Using `tt()` present the `output`.

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

...
```

:::
