---
title: "Data Visualization"
subtitle: "Week 4"
date: 2025-10-10
format: 
  html:
    embed-resources: true
toc: true
---


# Before we start

-   Any questions?


<center>
<a href="https://artur-baranov.github.io/nu-ps403-ds/scripts/ps403-d_04.qmd" class="btn btn-primary" role="button" download="ps403-d_04.qmd" style="width:200px" target="_blank">Download script</a>
</center>
<br>
<center>
<a href="https://artur-baranov.github.io/nu-ps403-ds/data/whogov.csv" class="btn btn-primary" role="button" download="whogov.csv" style="width:200px" target="_blank">Download data</a>
</center><br>


::: callout-note
## Review

First, load the `tidyverse` library

```{r}
#| message: false

library(tidyverse)
```

Load the World Values Survey data from the previous week's class.

```{r}
#| eval: false

wvs = ...
```

Rename the variables:

-   `B_COUNTRY_ALPHA` to `country`

-   `Q191` to `violence`

-   `Q194` to `political_violence`

```{r}
#| eval: false

...
```

Calculate the average and standard deviation of `political_violence` by `country` using `group_by()` and `summarize()`. Save that to a new `pviolence_mean` object. 

```{r}
#| eval: false

pviolence_mean = wvs %>%
  ... %>%
  summarize(mean = ...,
            sd = ...)
```

Present first observations in the `pviolence_mean` using `head()`.

```{r}
#| eval: false

head(pviolence_mean)
```

Leave only North American countries.

```{r}
#| eval: false

north_america = c("CAN", "USA", "MEX")

pviolence_mean_filtered = ...
```


Load the `tinytable` library and present a table describing North American countries using `tt()`. Round the numbers up to two digits. Change the background color of Mexico’s (MEX) standard-deviation cell to yellow. And if you have underscores or some other special symbols (for example, the variable has `_` underscores), don't forget to use `format_tt(escape = TRUE)`.

```{r}
#| eval: false

library(tinytable)

pviolence_mean_filtered %>%
  tt() %>%
  format_tt(...) %>%
  ...(i = 2,
      ...,
      background = "yellow")
```

And a small preview of what's going to come!

```{r}
#| eval: false

ggplot(wvs) +
  geom_point(aes(x = violence,
                 y = political_violence))
```


Now, clear the environment using the broom icon!


:::

# Agenda

-   Summarizing Data

-   Presenting Data with Graphs

-   Exploring Relationships Visually

# Exploring Data

Today we are working with WhoGov dataset, which provides information on Members of Cabinets. As usual, I recommend taking a look at their [codebook](https://politicscentre.nuffield.ox.ac.uk/media/4117/whogov_codebook.pdf).

```{r}
whogov = read.csv("data/WhoGov.csv")
```


First of all, these are the following variables we are going to work with today:

-   `country_name` is a country name

-   `n_individuals` number of unique persons in the cabinet

-   `leaderexperience_continuous` the number of years the person has been leader of the country in total.

-   `system_category` the regime type

-   `average_total` the average tenure for people in cabinet

Say, you are interested in the relationship between the leader's duration in the office (`leaderexperience_continuous`) and the tenure for people in cabinet (`average_total`).

# One Continuous Variable

Start with exploring the distribution of average tenure of people in the cabinet (`average_total`)

```{r}
ggplot(whogov) +
  geom_histogram(aes(x = average_total)) 
```

You can inform your reader by providing extra information. Add the average to the plot using `geom_vline()` and change the number of `bins` to 20. Save the graph to the `avgind_plot` object.

```{r}
#| warning: false

avgind_plot = ggplot(whogov) +
  geom_histogram(aes(x = average_total), bins = 20) +
  geom_vline(aes(xintercept = mean(average_total, na.rm = TRUE)), color = "#EE4B2B") 

avgind_plot
```

Let's name the axes and customize the graph. Feel free to explore ggplot themes [here](https://r-charts.com/ggplot2/themes).

```{r}
#| warning: false

avgind_plot = avgind_plot +
  labs(x = "Average Tenure ",
       y = "Count",
       title = "Distribution of Average Tenure",
       subtitle = "In Cabinets") +
  theme_bw()

avgind_plot
```

Change the breaks to a more readable format. 

```{r}
avgind_plot +
  scale_x_continuous(breaks = seq(0,                         # start sequence from
                                  max(whogov$average_total,
                                      na.rm = TRUE),         # end sequence with
                                  3))                        # increment
```

# Two Continuous Variables

Now, continue exploring the relationship we were interested in. Let's plot two distributions with densities. Make sure to go through the syntax. And be careful, now the X axis named incorrectly! 

-   Compare the distributions

-   What alternative plot would be more suitable for comparing distributions?

Add `xlim(-5, 20)`

```{r}
#| warning: false

ggplot(whogov) +
  geom_density(aes(x = average_total), fill = "red", alpha = 0.5) +
  geom_density(aes(x = leaderexperience_continuous), fill = "blue", alpha = 0.5) 
```

Let's draw a scatter plot using `geom_point()`. It's more appropriate if we are interested in a relationship, rather than distribution comparison. What's going on?

```{r}
ggplot(whogov) +
  geom_point(aes(x = leaderexperience_continuous,
                 y = average_total))
```

Let's zoom in to explore two political regimes. Use `system_category` to leave only `Military dictatorship` and `Parliamentary democracy`. 

```{r}
two_regimes = whogov %>% 
  filter(system_category %in% c("Military dictatorship",
                                                "Parliamentary democracy"))
```

Then, color the points.

```{r}
regime_tenure = ggplot(two_regimes) +
                geom_point(aes(x = leaderexperience_continuous,
                               y = average_total,
                               color = system_category))

regime_tenure
```

The patterns are seen clearer now! But let's also add panels of graphs with `facet_grid()`. Add `scales = "free"` argument to the facet grid function.

```{r}
#| warning: false 

regime_tenure = regime_tenure +
  facet_grid(system_category ~ .)

regime_tenure
```

Add title, name the axes and add a `theme_minimal()`.

```{r}
regime_tenure +
  labs(y = "Average Cabinet Tenure",
       x = "Leader's Tenure",
       title = "Leader's Tenure and Cabinet Tenure",
       color = "Political Regime") +
  theme_minimal()
```

Finally, let's see the correlation. Is it a big one?

```{r}
cor(two_regimes$n_individuals, two_regimes$leaderexperience_continuous)
```

For your reference, take a look on the [cheatsheet](https://rstudio.github.io/cheatsheets/html/data-visualization.html).

# Check list

- [ ] I have no trouble naming my axes and providing titles to the plot with `labs()` function.

- [ ] I know that I can draw distributions with `geom_density()`.

- [ ] I know I can work with `scale_x_...` and `scale_y_...` to change the presentation of X or Y axes.

- [ ] I can create separate plots for different groups using `facet_grid()` (and with `facet_wrap()` -- take a look on the exercises).

- [ ] I like customizing graphs with different themes! And I know how to change colors of elements in `ggplot()`.

# Extra

## Plotting a categorical variable

You can use histograms to count number of observations in each category, and then plot this count. Let's try this out with political regimes (`system_category`). First, count the number of observations in each group. This is a dataframe.

```{r}
regime_count = whogov %>%
  group_by(system_category) %>%
  count()

regime_count
```

Now, plot it using `geom_col()`.

```{r}
regime_count %>%
  ggplot(aes(x = system_category,
             y = n)) +
  geom_col() 
```

Change the order using `reorder()`, so the the count is presented consecutively.

```{r}
regime_count %>%
  ggplot(aes(x = reorder(system_category, n),
             y = n)) +
  geom_col() 
```

Now, customize the graph. Change the color of the bars with `fill = ` argument in `geom_col()`, flip the axes making it easier to read names of the regime with `coord_flip()`. Add names of the axes with `labs()`, and change the theme to `theme_dark`.

```{r}
regime_count %>%
  ggplot(aes(x = reorder(system_category, n),
             y = n)) +
  geom_col(fill = "lightblue") +
  coord_flip() +
  labs(x = "Political Rregime",
       y = "Count") +
  theme_dark()
```

## Plotting Trends

Let's explore average age of cabinets in the Switzerland. First, create the dataframe with data containing only the Switzerland data.

```{r}
swiss_tenure = whogov %>%
  filter(country_name == "Switzerland")
```

Now, plot it with `geom_line()`. Make `year` on X axis, and `average_total` on Y axis.

```{r}
ggplot(swiss_tenure,
       aes(x = year,
                y = average_total)) +
  geom_line() +
  geom_point()
```

## Plotting ovelapping plots

Imagine you are interested in comparing two distributions: `average_total` (the average tenure of the cabinet), and the `leaderexperience_continuous` (tenure of the leader). We have plotted it with in the beginning. 

```{r}
#| warning: false

ggplot(whogov) +
  geom_density(aes(x = average_total), fill = "blue", alpha = 0.5) +
  geom_density(aes(x = leaderexperience_continuous), fill = "red", alpha = 0.5) +
  xlim(-5,20)
```


If you want to compare two groups making one group appear on top of the other, you have an option of doing so using `ggridges`. But you also need to restructure the data to a **long format**.

```{r}
library(ggridges)

whogov_long = whogov %>%
  select(average_total,
         leaderexperience_continuous) %>% # choose the variables
  pivot_longer(1:2)                       # make the data of a longer format

head(whogov_long)
```

Finally, plot the restrucutred data.

```{r}
ggplot(whogov_long) +
  geom_density_ridges2(aes(x = value,
                           y = name)) +
  xlim(-5, 20)
```



# Optional Exercises

Explore `n_individuals` variable. Draw a histogram, 

```{r}
#| eval: false

...
```

Explore a new function. Use `facet_wrap()` with `system_cateogry`. Compare it with `facet_grid()`. What's the difference?

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

...
```

:::

Make the `scale = "free"` in `face_wrap()`.

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

...
```

:::

Now, explore relationship between `leaderexperience_continuous` (X) and `n_individuals` (Y). Draw a `geom_point()`

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

...
```

:::

Add a `facet_wrap()` layer to the previous graph, by `system_category`.

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

...
```

:::

Customize the graph. Name the axes, make the scales free, add color to the observations dots on the graph based on the `system_category`.

::: {.callout-tip icon="false"}
## Solution

```{r}
#| eval: false

...
```

:::
