---
title: "Data Presentation"
subtitle: "Week 3"
date: 2025-10-03
format: 
  html:
    embed-resources: true
toc: true
---

# Before we start

-   Any questions?

<center>
<a href="https://artur-baranov.github.io/nu-ps403-ds/scripts/ps403-d_03.qmd" class="btn btn-primary" role="button" download="ps403-d_03.qmd" style="width:200px" target="_blank">Download script</a>
</center>
<br>
<center>
<a href="https://artur-baranov.github.io/nu-ps403-ds/data/cpds.xlsx" class="btn btn-primary" role="button" download="cpds.xlsx" style="width:200px" target="_blank">Download data</a>
</center><br>

::: callout-note
## Review

First, load the `tidyverse` library

```{r}
#| message: false

library(tidyverse)
```

Then, load the V-Dem data

```{r}
#| eval: false

load(url("https://github.com/vdeminstitute/vdemdata/raw/6bee8e170578fe8ccdc1414ae239c5e870996bc0/data/vdem.RData"))
```


You are interested in corruption in various geographical regions. Select year (`year`), region (`e_regionpol_7C`) and corruption (`v2x_corr`) variables from the `vdem` dataset.

```{r}
#| eval: false

corruption_data = ... %>%
  ...(year, ..., v2x_corr) 
```


As we are working with the V-Dem data, let's rename variables so it's more straightforward.

```{r}
#| eval: false

corruption_data = corruption_data ...
  ...(region = e_regionpol_7C,
      corruption = ...)
```

Calculate the average of the `corruption` variable. Don't forget about the `na.rm = TURE` argument!

```{r}
#| eval: false

...
```

Using `group_by()` and `summarize()` calculate the average corruption for all regions in the dataset.

```{r}
#| eval: false

corruption_data ...
  (...) %>%
  ...(average = ...(..., na.rm = T))
```

Lastly, let's see the distributions of corruption across regions. Draw a boxplot below with `region` variable on the X axis, and `corruption` variable on the Y axis.

```{r}
#| eval: false

ggplot(data = corruption_data) +
  ...(aes(x = ..., y = ...))
```

Something is wrong, right? We haven't checked the classes of the variables, and apparently the `region` variable has to be changed. Let's check it's class first.

```{r}
#| eval: false

...(corruption_data$...)
```

What class should it be? Let's recode it directly on the plot.

```{r}
#| eval: false

...(...) +
  geom_boxplot(aes(x = ...(region), y = ...))
```

:::

visualizations

one variable
+ variance/covariance
+ sd
+ mean

two variables
correlation?
geom_point

tinytable



