---
title: "Publishing and Quarter Review"
subtitle: "Week 9"
date: 2025-11-14
format: 
  html:
    embed-resources: true
toc: true
---

# Before we start

-   No Discussion Section and Office Hours Next Week (but available by demand)

-   Any questions?

# Agenda

-   Quick Review

-   Example of Quarto and Overleaf for Publishing

# Quarto for Publishing

Quarto and Overleaf are both excellent tools for publishing. For this class, it's better to stick with Quarto -- the scope of your projects makes it a suitable choice. However, if your papers become longer or you plan to collaborate with others, Overleaf is the better option.

Here is how to get started.

-   [Paper Template for Quarto](https://github.com/gustavo-diaz/ps403/tree/main/paper_template). 
  
-   [Overleaf](https://www.overleaf.com). Click `New project` -> `Example project` to get started.

Helpful tips:

-   [GitHub Desktop Tutorial](https://joshuadull.github.io/GitHub-Desktop/02-getting-started/index.html)

-   [Exporting .bib file from Zotero](https://libguides.usask.ca/c.php?g=218034&p=1446406)

-   Google Scholar has BibTeX citations!


<center>
<a href="https://artur-baranov.github.io/nu-ps403-ds/scripts/ps403-d_09.qmd" class="btn btn-primary" role="button" download="ps403-d_08.qmd" style="width:200px" target="_blank">Download script</a>
</center>


::: callout-note
## Review

First of all, load the usual libraries and the Transnational Justice dataset from the previous week.

```{r}
#| message: false
library(tidyverse)
library(marginaleffects)
library(GGally)

tjet = read.csv("data/tjet.csv")
```

Let's explain under what circumstances countries adopt reparations policies. Could the involvement of international organizations influence a state's decision to implement such policies?

-   `rep_binary` presence of reparations policy

-   `ICC_investigation` presence of ICC investigation

-   `uninv` ongoing UN investigations, count

-   `trials_domestic`	count of domestic trials

First, let's see if the probability of reparations (`rep_binary`) is increasing when ICC intervenes (`ICC_investigation`). Set up a `glm()` logit model.

```{r}
#| eval: false

model_icc = ...
```

Present summary of the model.

```{r}
#| eval: false

summary(model_icc)
```

Now, let's draw the graph using `plot_predictions()` from `marginaleffects` library. What is worrisome? 

```{r}
#| eval: false

...(model_icc, 
                 ... = c("ICC_investigation"))
```

Now, let's add a couple of variables. Add `uninv` and `trials_domestic`.

```{r}
#| eval: false

model_full = glm(rep_binary ~ ICC_investigation + ..., family = binomial, data = tjet) 
```

Present `plot_predictions()` for `ICC_investigation` only.

```{r}
#| eval: false

...
```

Finally, present the `modelsummary()` table for both models. Compare.

```{r}
#| eval: false

modelsummary(list("ICC Model" = model_icc,
                  "Full Model" = model_full),
                   stars = c("*" = 0.05),
                   gof_omit = "AIC|BIC|Log.Lik|F|RMSE" ,
                   statistic = "p.value")
```

Don't forget to clean the environment.

:::


# Exploring Data

Today we are working with World Happiness Report data, the one we had on the [week 5](https://artur-baranov.github.io/nu-ps403-ds/ps403-d_05.html).

-   `Country_name` is the name of the country

-   `Ladder_score` is the happiness score

-   `Continent` 

Load the data.

```{r}
whr = read.csv("data/WHR.csv")
```

Explore the following variables with the `ggpairs()` function from `GGally`. What can you see on the graph?

```{r}
#| message: false

library(GGally)
ggpairs(whr,
        columns = c("Ladder_score",
                    "Logged_GDP_per_capita",
                    "Continent"))
```

Let's zoom in the Happiness Score by Continent. Is Africa different from Europe in terms of the Happiness?

```{r}
ggplot(whr) +
  geom_boxplot(aes(x = Ladder_score, y = Continent)) +
  labs(x = "Happiness") +
  theme_bw()
```

Zoom in even more. Use `filter()` to leave only the continents we are analyzing. And leave only the variables we want to work with: `Country_name`, `Ladder_score`, `Continent` and `Logged_GDP_per_capita`.

```{r}
ae_whr = whr %>%
  filter(Continent %in% c("Africa", "Europe")) %>%
  select(Country_name, Ladder_score, Continent, Logged_GDP_per_capita)
```

Plot again. Make color more transparent with `alpha = 0.5`.

```{r}
ggplot(ae_whr) +
  geom_boxplot(aes(x = Ladder_score, y = Continent, fill = Continent), alpha = 0.5) +
  labs(x = "Happiness",
       y = NULL) +
  theme_light()
```

# Formal Tests

Finally, let's test it. Interpret the results.

```{r}
t.test(ae_whr$Ladder_score ~ ae_whr$Continent)
```

We can confirm it with the simple model. Estimate the Happiness Score depending on the continent, and plot the confidence intervals for the estimate. Is it different from the boxplots?

```{r}
lm(Ladder_score ~ Continent, 
   ae_whr) %>%
  plot_predictions(by = "Continent") +
  coord_flip()
```

Now, let's plug in the Wealth, measured as Log GDP per capita in the model. Explore difference.

```{r}
model = lm(Ladder_score ~ Logged_GDP_per_capita + Continent, ae_whr)
plot_predictions(model,
                 condition = c("Logged_GDP_per_capita", "Continent"))
```

Present summary. Do `Continent` variable matter?

```{r}
summary(model)
```

Let's add one more variable, and run the model on the whole dataset. What does it tell us?

```{r}
lm(Ladder_score ~ Social_support + Logged_GDP_per_capita + Continent, whr) %>%
  summary()
```


# What's next?

Congrats with finalizing the class! No optional exercises this time.

-   [Helpful resource for R](https://rstudio.github.io/cheatsheets/)

-   Check the [main page](https://artur-baranov.github.io/nu-ps403-ds/) for additional resources

-   Learning to code is like learning a new language: if you're interested, keep practicing!

-   Linear models next!

# What we have covered in coding

| Library         | Functions                          | Description                                    |
|-----------------|------------------------------------|------------------------------------------------|
| tidyverse       | `filter()`, `mutate()`, `ggplot()` | data wrangling and visualization               |
| modelsummary    | `modelsummary()`                   | present good looking model summary tables      |
| broom           | `tidy()`                           | extract additional information from the models |
| marginaleffects | `plot_predictions()`               | calculate and visualize marginal effects       |
| GGally          | `ggpairs()`                        | extension to ggplot                            |
| tinytable       | `tt()`                             | present good looking tables                    |
| patchwork       | `/`, `+`                           | plot several graphs together                   |

# Datasets we have used

| Dataset                                           | Description                               | Link                                                     |
|---------------------------------------------------|-------------------------------------------|----------------------------------------------------------|
| **V-Dem**                                         | Measures democracy worldwide              | [V-Dem](https://www.v-dem.net/)                          |
| **World Happiness Report**                        | Annual happiness report                   | [World Happiness Report](https://worldhappiness.report/) |
| **Who Governs**                                   | Dataset on political elites               | [Who Governs](https://doi.org/10.1017/S0003055420000490) |
| **SIPRI**                                         | Data on military operations               | [SIPRI](https://www.sipri.org/databases)                 |
| **Comparative Political Data Set**                | A dataset covering political institutions | [CPDS](https://www.cpds-data.org/)                       |
| **World Values Survey**                           | Survey measuring values across the globe  | [WVS](https://www.worldvaluessurvey.org/wvs.jsp)         |
| **Transitional Justice Evaluation Tools Dataset** | Dataset assessing transitional justice    | [TJET Dataset](https://transitionaljusticedata.org)      |


# Check List

- [ ] I know that every lab has check list below! And I will use it to navigate what we have learned

- [ ] R doesn't scare me anymore

- [ ] I have developed the intuition behind application of quantitative methods and happy to learn more!
