---
title: "Logistic Regression"
subtitle: "Week 8"
date: 2025-11-07
format: 
  html:
    embed-resources: true
toc: true
---

# Before we start

-   Any questions?


<center>
<a href="https://artur-baranov.github.io/nu-ps403-ds/scripts/ps403-d_08.qmd" class="btn btn-primary" role="button" download="ps403-d_08.qmd" style="width:200px" target="_blank">Download script</a>
</center>
<br>
<center>
<a href="https://artur-baranov.github.io/nu-ps403-ds/data/..." class="btn btn-primary" role="button" download="..." style="width:200px" target="_blank">Download data</a>
</center><br>


::: callout-note
## Review

Load the `tidyverse`, `marginaleffects`, and `GGally` libraries and sipri dataset

```{r}
#| message: false
library(tidyverse)
library(marginaleffects)
library(GGally)

sipri = read.csv("data/sipri.csv")
```

Using `GGally` library, visualize the relationship between `Import` and `Regime` variables

```{r}
ggpairs(sipri, c("Import", "Regime"))
```

Set up a linear model. Predict `Import` by `Regime`. Visualize the model with `plot_predictions()` from `marginaleffects`.

```{r}
model1 = lm(Import ~ Regime, sipri)


plot_predictions(model,
                 by = "Regime")
```

What is the reference category in this case?

```{r}

```

Using `relevel()`, set the reference category to `Minimally Democratic`

```{r}

```

Set-up the model again, and plot the `ggpredict()`. Did anything change?

```{r}

```

Extract the model information using `tidy()` function from `broom` library.

```{r}

```

Visualize the effects plot. Is it different to the plot generated by `ggpredict()`?

```{r}

```

:::

# Agenda

logistic regression +
interpretations and marginal effects

bootstrap standard error from rsample from logistic regression hw7 - 4.3 

Today we are working with Transitional Justice Evaluation Tools dataset. You can explore their codebook [here](https://transitionaljusticedata.org/en/downloads.html). Let's load the data.

```{r}
#| eval: false

tjet = read.csv("data/tjet.csv")
```


We are going to use the following variables:

-   `country` and `year`

-   `dtr`	democratic transition

-   `tcs_operated`	truth commission operated

-   `trials_domestic`	count of domestic trials per country and start year

-   `region`	United Nations region

Subset the above variables with `select()`

```{r}
#| eval: false

tjet_sub = ... %>%
  ...(country, year, dtr, tcs_operated, trials_domestic, region) 
```

Leave more recent observations. Say, starting year 1975. 

```{r}
#| eval: false

tjet_sub = tjet_sub %>%
  ...(year >= 1975)
```

Let's briefly try it out!

```{r}
#| eval: false

model_glm = glm(dtr ~ trials_domestic + tcs_operated, tjet_sub, family = binomial(link = "logit")) 
summary(model_glm)
```

Now, visualize how it looks like. Usually, we would expect to see slightly more "S" shaped curve. But still - good! 

```{r}
#| eval: false

ggpredict(model_glm, term = c("trials_domestic", "tcs_operated")) %>%
  plot(show_data = T)
```

Compare to linear probability model **without** country and year fixed effects.

```{r}
#| eval: false

ggpredict(model, term = c("trials_domestic", "tcs_operated")) %>%
  plot(show_data = T)
```

Present both: linear probability model without fixed effects and logistic regression. The problem, though, that Logit is way harder to interpret! Stick to marginal effects!

```{r}
#| eval: false
library(modelsummary)

modelsummary(list("LPM" = model,
                  "Logit" = model_glm),
             stars = T)
```


A final note, in most cases you would expect to see logistic regression as something like

```{r}
#| warning: false

glm(am ~ hp + wt, data = mtcars, family = binomial) %>%
  ggpredict(terms = c("hp")) %>%
  plot(show_data = T)
```